Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-02-20 agujournaltemplate.tex
Focus Phase Day 4 - Intro Progress

    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}. These communities consist of a diverse set of species that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014,henson_observing_2016}. 
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin off the coast of Venezuela between 1995 and 2017 \cite{muller-karger_scientific_2019}. The Cariaco basin is one of the most productive ecosystems in the Caribbean Sea \cite{bringue_physico-chemical_2018}, with seasonal upwelling driven by trade winds during December through May and a secondary upwelling observed in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. The basin is anoxic at depths below \qty{\sim 275}{m}, due to microbial respiration of sinking organic matter and restricted bottom water circulation, leaving a rich sedimentary record for the study of past climate \cite{hughen1996nature}.

@@

    "The Cariaco Basin, located ...." - general hydrography/depth, etc.
        The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. 
        A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{astor_seasonal_2003}. 
        The basin is effectively separated from the Caribbean Sea by a sill with a mean depth of about 100 m. The water below the sill is poorly ventilated, which, combined with the high surface productivity, results in permanent anoxic conditions at depths below 275 m \cite{thunell_organic_2000}.

        Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.

@@

- We are not able to fully resolve....

@@

% Pinckney et al. 2015:
- major thing is that previous paper only looked at microscopy to 55m (Mutshinda and Pinckney), but now we look at full 100 m!
%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014). Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014). This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012). Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013). During this period, a collapse in the sardine industry also occurred. The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012). These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
% Consistent with the results of Taylor et al. (2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m. However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3). We argue that the reduction in cell numbers is likely only true of larger cells. Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000). Furthermore, Mousing et al. (2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability. Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e. below 56 m). In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years. In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014). Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.

>>>>

    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}. These communities consist of a diverse set of species that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.

    %DIVERSITY PARAGRAPH HERE? yes...

    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test-bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)

    Since the inception of the time series, the Cariaco basin has undergone marked shifts...

    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts and high performance liquid chromatography (HPLC) pigment analysis. Of these, all cover the entire time series except the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.

    - Taylor 2012 ()
    \citeA{taylor_ecosystem_2012} analyzed data up until 2010 and found significant decadal trends of sea surface temperature (SST) increase, upwelling intensity decrease and diminished phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified via microscopy showed a shift towards smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass. The authors explained the change in physical regime in the Cariaco basin with a north-easterly shift of the Inter-Tropical Convergence Zone (ITCZ) and a resulting weakening trend in Trade Winds.

    - Mutshinda 2013 ()
    Using a hierarchical Bayesian modeling approach, changes in dominant species quantified via microscopy cell counts between 1995 and 2011 were found to be best predicted by changes in temperature, pH and irradiance, and less so by salinity or macronutrient concentrations \cite{mutshinda_environmental_2013}. In particular, higher temperatures were associated with a lower abundance of diatoms.

    - Irwin 2015 ()
    With the same dataset of microscopy cell counts and a MaxEnt statistical modeling approach, \citeA{irwin_phytoplankton_2015} calculated realized niches for the dominant species for SST, irradiance and nitrate concentration. As trends in the realized niches were observed over the course of the time series, this seemed to show the adaption of species to climate change effects.

    - Pinckney 2015 ()
    The phytoplankton community was investigated in detail by \citeA{pinckney_phytoplankton_2015}, using the microscopy counts and HPLC data from the time ranges 1996 to 2000 and 2006 to 2010. Between these two time periods, they observed that the drastic reduction in diatoms counted using microscopy did not correspond to a similar reduction in diatom abundance calculated from the pigment composition, which is most likely explained by a shift to diatom species with smaller cell sizes. Additionally, the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling  and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, the shifts in the phytoplankton community did not result in a significant change to the export of particulate organic carbon (POC) to depth as measured by sediment traps deployed during the time series.

    - Lorenzoni 2015 ()
    \citeA{lorenzoni_characterization_2015} focused their analysis on HPLC data retrieved between 2006 and 2012 and related seasonal changes in the pigment composition to spectral absorption from remote sensing data and concentrations of organic carbon. They observed a high seasonal variability, with microphytoplankton (mostly diatoms) dominating upwelling periods and picophytoplankton (mostly cyanobacteria) dominating during the rainy season. These community compositions were also reflected in the POC:TChla ratio, which was higher for diatom dominated assemblages and lower for those dominated by cyanobacteria. 
    % "The difference between the taxonomic data and the pigment data suggests that the microscopy taxonomy may not capture the variability in the pico-phytoplankton. Since the beginning of the time-series in 1995, there has been a change in the abundance of phytoplankton size classes based on diagnostic pigments: micro-phytoplankton has decreased year-round, while picophytoplankton has increased, particularly during the rainy season. This is consistent with the decrease in larger phytoplankton previously reported for the region." Lorenzoni et al. 2015

    - AMO (ref to 2019 Review)
    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al. 2015). These climatic changes had a profound impact throughout the Atlantic (Alexander et al. 2014, Nye et al. 2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.

    - Finally what it says in Review about Phytoplankton community
    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

    % Hone in on Diversity Aspects, and what has been done there?
    - Highlight diversity aspect of Pinckney
    - Highlight Mutshinda etc. How they used Niches/Phyto data
    - What are the questions in Diversity TS data? -> Esteban's papers

    % Finally say what WE are going to do! What are the open questions, what is new?

@@

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below \qty{\sim 200}{m} depth.

@@

- We are not able to fully resolve....
% "Phytoplankton communities exhibit a great deal of variation due to a combination of regulation of growth by bottom-up factors such as resource availability (Grover 1997), variation in mortality due to grazing, viral and parasitoid attack (Alpine & Cloern 1992; Mu ̈ hling et al. 2005), and demographic drift among species (neutral variation). Bottom-up factors can be complex, depending on nutrient supply rates and ratios (Tilman 1982), light availability and photoinhibition (Six et al. 2007; Alderkamp et al. 2010), species-specific competitive ability for resources (Litchman 2007) and other factors influencing physiological responses such as temperature (Eppley 1972). Variation in mortality is much less frequently measured than the factors promoting growth due to the technical challenges in making measurements. None of our environmental variables are good proxies of grazing rates or whether the abundance of a particular population is determined by grazer control or resource supply at a particular time. We have ident" from Mutshinda et al. 2013

%"Taylor et al. 2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys.. This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity. A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO

@@

% Pinckney et al. 2015:
- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014). Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014). This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012). Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013). During this period, a collapse in the sardine industry also occurred. The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012). These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
% Consistent with the results of Taylor et al. (2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m. However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3). We argue that the reduction in cell numbers is likely only true of larger cells. Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000). Furthermore, Mousing et al. (2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability. Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e. below 56 m). In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years. In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014). Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.
